---
title: "a05:Data Visualization: ggplot2"
author: "Victor Knowles"
date: "2026-03-16"
---
#Introduction: This report visualizes visitor activity at restoration locations in the Sacramento-San Joaquin Delta. The goal of this analysis is to explore patterns in visitor types across different restoration sites and visualize their geographic locations.
 
#Overview
```{r}
library (readr)
library(dplyr)
library(tidyr)
library(forcats) # makes working with factors easier
library(ggplot2)
library(leaflet) # interactive maps
library(DT) # interactive tables
library(scales) # scale functions for visualization
library(janitor) # expedite cleaning and exploring data
library(viridis) # colorblind friendly color palette
```

#Getting the data ready
```{r}
# read in data from the data directory after manually downloading data, published 8/6/2020
#Citation:
#Kraus-Polk, A. and B. Milligan. 2020. Sacramento-San Joaquin Delta Socioecological Monitoring ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/e5bba3945d7f94c3005a7f8c38d00802 (Accessed 2026-04-07).

delta_visits_raw <- read_csv(here::here("data/raw data/Socioecological_monitoring_data.csv"))
```
```{r}
delta_visits <- delta_visits_raw %>%
  janitor::clean_names()
```

```{r}
visits_long <- delta_visits %>%
  pivot_longer(cols = c(sm_boat, med_boat, lrg_boat,
                        bank_angler, scientist, cars),
               names_to = "visitor_type",
               values_to = "quantity") %>%
  rename(restore_loc = eco_restore_approximate_location) %>%
  select(-notes)
```

```{r}
daily_visits_loc <- visits_long %>%
  group_by(restore_loc, date, visitor_type) %>%
  summarise(daily_visits = sum(quantity))
```

#Plotting with ggplot2
```{r}
# Line 165 in wrangling file
# use cleaned and summarized dataset
ggplot(data = daily_visits_loc, 
       
# Map variables to axes and colors
aes(
  y = restore_loc,      # restoration locations on y-axis
  x = daily_visits,     # number of visits on x-axis
  fill = visitor_type   # color bars by visitor type
) 
) +
  
  #Draw bar chart
    geom_col() +
    labs(x = "Number of Visits", # label x-axis
         y = "Restoration Location", # label y-axis
         fill = "Type of Visitor", # legend title
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
  
  # Apply clean theme
    theme_bw() +
  
  # Adjust legend and remove axis ticks
    theme(legend.position = "bottom", # move legend to bottom
          axis.ticks.y = element_blank()) # remove y-axis ticks
```
#Plot description: The following plot shows the total number of visit to each restoration location, grouped by visitor type. This visualization helps compare how different visitor groups interact with restoration areas.

#Analysis: The bar plot shows that visitor activity varies across restoration locations, with some sites receiving substantially higher numbers of visits than others. Different visitor types also contribute unevenly to total visitation, with certain categories such as boats or anglers appearing more frequently at specific locations. This suggests that restoration areas are used diffently depending on both location and visitor type.

```{r}
# Create dataset of unique restoration locations
locations <- visits_long %>%
  distinct(restore_loc, .keep_all = TRUE) %>%   # keep one row per location
  select(restore_loc, latitude, longitude)      # keep only relevant columns

# Create interactive map
leaflet(locations) %>%

  # Add satellite imagery basemap
  addProviderTiles(providers$Esri.WorldImagery) %>%

  # Add label overlay
  addProviderTiles(providers$CartoDB.PositronOnlyLabels) %>%

  # Add circular markers for each location
  addCircleMarkers(
    lng = ~longitude,        # longitude (x-coordinate)
    lat = ~latitude,         # latitude (y-coordinate)
    popup = ~restore_loc,    # label shown when clicked
    radius = 5,              # marker size

    # Fill properties
    fillColor = "salmon",    # marker color
    fillOpacity = 1,         # marker opacity

    # Border properties
    stroke = TRUE,           # enable border
    weight = 0.5,            # border thickness
    color = "white",         # border color
    opacity = 1              # border opacity
  )
```
#Map description: The map below displays the geographic locations of the restoration sites included in the dataset.Each marker represents a restoration location in the Sacramento-San Joaquin Delta.

#Analysis: The map provides spatial context for these patterns by showing where each restoration site is located within the Delta region. Clusters of point show the areas with multiple restoration sites, which may correspond to regions of higher ecological activity or accessibility.

#Conclusion: Overall, the data plot and map highlight both the variation in visitor activity and the geographic distribution of restoration sites in the Delta. Together, the plot and map provide insight into how human activity is distributed across restoration areas, which could be useful for management and conservation planning.
